---
title: "Approximating Quantities and Probabilities from The Wright-Fisher Model"
output: html_notebook
---

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Introduction

The Wright-Fisher model provides a way of modelling the random inheritance of
genes from one generation to the next.  In its simplest incarnation, the population size
is constant each generation: $N$ haploid individuals.  Each individual can thus be
represented by a single gene copy.  The gene copies in generation $t+1$ are obtained
by sampling _with replacement_ from amongst the gene copies in generation $t$.

Let us assume we are concerned with a single locus at which there are two different
alleles, $A$ and $a$.  And we let the random variable $X_t$ record the number of
$A$ alleles (out of $N$) in the population at generation $t$.  Initially (at $t=0$)
there are $x_0$ copies of the $A$ allele. Under the Wright-Fisher model's
sampling with replacement from the previous generation, we have a binomial distribution with
$N$ draws, each with a success probability $p$ of $x_t/N$
$$
X_{t+1}  \sim \mathrm{Binomial}(N; x_t/N)~~~~,~~~~t = 1, 2, \ldots
$$

Given this setup, we can write a function to simulate $R$ different replicates of the Wright-Fisher
model, starting from an initial allele frequency of $X_0$, and proceeding $T$ generations until $X_T$
is obtained.  I have provided such a function that returns the results for all the replicate simulations
in a tibble with three columns:

- **rep** the number of the replicate run
- **t** the number of the generation
- **x** the number of copies of the $A$ allele at generation $t$ in from replicate `rep`.

If you are not  familiar with a `tibble` it is similar to a data frame, but it forms
the basis for many operations in the "tidyverse" way of using R, made popular by 
Hadley Wickham.  The tidyverse provides a unified way of handling many different
genres of data using ideas that are foundational to relational databases. There should
be at least a few individuals in each break-out group that are familiar with the tidyverse
that can help those that are not experienced with it. If you have not run across the
tidyverse before, then it would be worth reading up on it.  An excellent book
that covers many aspects of the tidyverse
is available at [https://r4ds.had.co.nz/](https://r4ds.had.co.nz/).

Here is the function:
```{r}
#' simulate R reps of a Wright-Fisher model of size N haploids for T generations from an initial allele count of X0
#'
#' Default values for the simulatiions are given as shown.
#'
#' @param R number of replicate simulations to do
#' @param N number of haploids, each generation, in the population
#' @param T number of generations of sampling to simulate
#' @param X0 the initial number of A alleles.  Must be <= N
WF <- function(
  R = 1000,
  N = 100,
  T = 20,
  X0 = 30
) {
  stopifnot(X0 <= N)
  
  lapply(1:R, function(r) { # cycling over the different replicates
    # initialize a vector to store the results
    x <- rep(NA, T)
    
    # get X1
    x[1] <- rbinom(n = 1, size = N, prob = X0 / N)
    
    # cycle over the remaining generations (for loop is slow but makes
    # it easier to read what is going on)
    for (t in 2:T) {
      x[t] <- rbinom(n = 1, size = N, prob = x[t - 1] / N)
    }
    
    # return a tibble of results
    tibble(
      rep = r,
      t = 1:T,
      x = x
    )
  }) %>%
    bind_rows()   # at the end, bind them all into a single tibble
}
```

It is not super fast, but it works well enough. It also returns every $X_t$ for every replicate
run of the Wright-Fisher model. So we can estimate a variety of quantities with the output, using
Monte Carlo.

## Examples

To recreate the histogram of $x_{10}$ values in the lecture you can do:
```{r}
set.seed(5)
ex1 <- WF(
  R = 1e4,
  N = 100,
  T = 12,
  X0 = 30
  )

# then count up the different x_10 values
counts10 <- ex1 %>%
  filter(t == 10) %>%
  count(x) %>%
  mutate(prob = n / sum(n))

# and, plot it
g <- ggplot(counts10, aes(x = x, y = prob)) + 
  geom_col(fill = "white", colour = "black", size = 0.3) +
  theme_bw()

g
```

Or to look at the allele frequencies over each generation from 1 to 12, we could do:
```{r}
counts_all <- ex1 %>%
  count(t, x)

# and, plot it
ggplot(counts_all, aes(x = x, y = n)) + 
  geom_col(fill = "white", colour = "black", size = 0.3) +
  theme_bw() +
  facet_wrap(~ t, ncol = 3)
```

## Exercises

Now that you know how to simulate from the Wright-Fisher model using the `WF()` function,
here are three exercises:

1. Use Monte Carlo to approximate the mean number of $A$ alleles at generations 1 through 20,
for a haploid population of size $N = 100$, starting with 30 $A$ alleles.
For each one, use 1000 replicates.
```{r}
S <- WF(
  R = 1e4,
  N = 100,
  T = 20,
  X0 = 30
  )
S %>%
  group_by(t) %>%
  summarise(meanX_t = mean(x))
```

2. As you should have verified through simulation above, the expected allele frequency
remains unchanged, over time, in a Wright-Fisher model.  However, the expected heterozygosity
decreases.  That is, the probability that two gene copies, sampled at random from the population
are different alleles, decreases over time, on average.  At time $t$ in a replicate, the
probability of drawing two gene copies of different alleles is, in our case, equal to
$$
H = 2\frac{X_t}{N}\frac{N-X_t}{N-1}
$$
Use the results from the simulated W-F population in exercise #1 to approximate the
heterozygosity at generations 1--20.
```{r}
S %>%
  group_by(t) %>%
  summarise(meanH_t = mean( (x * (100 - x)) / (100 * 99)))
```

3. Simulate 1,000 replicates of a haploid population of size 100 for 1,000 generations,
with $X_0 = 30$, and, from the output, approximate (via histogram) the distribution of time until fixation
(i.e., the first generation $t$, at which $X_t = 0$ or $X_t = 100$.)
```{r}
S2 <- WF(
  R = 1000,
  N = 100,
  T = 1000,
  X0 = 30
)

fix_times <- S2 %>%
  arrange(rep, t) %>% # just to ensure things are sorted correctly
  group_by(rep) %>%
  filter(x == 0 | x == 100) %>% # only keep the time steps after the allele is fixed
  slice(1) # take the first of those times in which the allele is fixed

ggplot(fix_times, aes(x = t)) +
  geom_histogram(binwidth = 1)
```

When doing MCMC you might often work like this: a large number of variable states get
written to disk or stored in a variable, and then after the fact you can use them in
Monte Carlo.