---
title: "Bonus excercise: use Monte Carlo to approximate a definite integral -- KEY"
output: html_notebook
---


```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
```


Consider a somewhat nasty-looking curve in two dimensions.  We
define $z$ to be a function of $x$ and $y$ that looks like this: 
$$
z = f(x, y) = \biggl[1 + e ^{-|x + 2y|} \cos\biggl(0.2 \sqrt{x^2 + y^2}\biggr) \biggr] \times I\{-10 < x,y < 10 \}
$$
In other words, it is a function that returns 0 everywhere outside of a square
of width 20, centered on the origin.  But within that square it returns
positive values.

Here is some code that defines it:
```{r}
ugly <- function(x, y) {
  ifelse(x < -10 | y > 10 | x > 10 | y < -10, 
         0,
         1 + exp(-abs(x + 2 * y) / 10) * cos(0.2 * sqrt(x^2 * y^2))
  )
}
```

And, we can use that to visualize the surface by evaluating a lot of points and
plotting it with the plotly package.  Go ahead and rotate and manipulate the thing.
```{r, fig.width=8}
vals <- seq(-12, 12, by = 0.05)
pts <- expand_grid(
  x = vals,
  y = vals
) %>%
  mutate(z = ugly(x, y))

zmat <- matrix(pts$z, ncol = length(vals), nrow = length(vals))

plot_ly(x = vals, y = vals, z = ~zmat) %>%
  add_surface()
```


The volume under that curve is given by the definite integral:
$$
\int_{-10}^{10} \int_{-10}^{10} f(x, y) dxdy = \int_{-10}^{10} \int_{-10}^{10} \biggl[1 + e ^{-|x + 2y|} \cos\biggl(0.2 \sqrt{x^2 + y^2}\biggr) \biggr] dxdy
$$
That looks a little hairy.  It may have an analytical solution, or it could
be evaluated numerically with high accuracy.  But, if you
rewrite that integral as an expection over a two-dimensional uniform variable on the square from -10 to 10 in $x$ and also from -10 to 10 in $y$, then you get:
$$
20 \times 20 \times \int_{-10}^{10} \int_{-10}^{10} f(x, y) \frac{1}{20}\frac{1}{20}dxdy =
400 \times \mathbb{E}[f(x, y)]
$$
where the expectation is taken over a 2-D uniform variable from -10 to 10 in both x and y.

Given that as a preamble, your mission it to approximate that integral using Monte Carlo sampling.

Good luck!  The answer is $\approx 423.7$.

Here is how that can be done:
```{r}
set.seed(123)
xs <- runif(1e6, -10, 10)
ys <- runif(1e6, -10, 10)

answer <-  400 * mean(ugly(xs, ys))

answer
```
